From physical principles to relativistic classical Hamiltonian and Lagrangian particle mechanics

### 2015-03-15 2014CompleteClassicalMechanics/completeClassicalMechanics.tex
Acknowledgements, bibliography and other minor edits

Here we are simply expressing $\omega$, $v$ and $w$ with their respective components and underlining the fact that $\omega$ defines the metric conserved under canonical transformations.

@@

\begin{defn}\label{inertialMass}
Let $p_\alpha$ be a component of conjugate momentum and $u_\alpha \equiv g_{\alpha \beta}u^\beta$ be a contravariant component of the four velocity. We define the \emph{inertial mass} as the cardinality conversion constant $m \equiv \mathfrak{C}^p_u$ between $p$ and $u$. That is, if $dp_\alpha=mdu_\alpha$ then $n^p(dp_\alpha)=n^u(du_\alpha)$.\footnote{Note that \ref{kinematicAssumption} rules out massless particles: they can have the same position/velocity with different momentum, therefore the same trajectory for different states. When generalizing to field theory, the derivative of the position will be replaced by the derivative of the field (i.e. its oscillation in space-time), which holds for massive and massless fields.}
\end{defn}

@@

Given \ref{initialConditions}, there must exist $p_\alpha=p_\alpha(x^\beta , u^\gamma)$. We express it in terms of $u_\alpha\equiv g_{\alpha \beta} u^\beta$. We have:
\begin{align*}
\omega &= q^i\wedge p_i - ct \wedge E/c \\
&=x^\alpha \wedge p_\alpha \\
&=dx^\alpha \wedge \frac{\partial p_\alpha}{\partial u_\beta}du_\beta + dx^\alpha \wedge \frac{\partial p_\alpha}{\partial x^\gamma}dx^\gamma \\
&=\frac{\partial p_\alpha}{\partial u_\beta}du_\beta dx^\alpha \\
\end{align*}
Consider the expression $m du_\alpha dx^\alpha$: this invariant gives us the density of labels in position and velocity converted to the conjugate variables. That is: $\omega=m du_\alpha dx^\alpha$. Combining the two:
\begin{align*}
\frac{\partial p_\alpha}{\partial u_\beta} &= m \delta^\beta_\alpha \\
p_\alpha &= m g_{\alpha \beta}u^\beta + \hat{p}_\alpha(x^\gamma)
\end{align*}
where $\hat{p}$ is an arbitrary function.\footnote{The conjugate momentum $p$ and the kinetic momentum $mu$ are therefore different in general: only their differentials are linked. Moreover: $p$ is unique only once $\hat{p}$ is fixed, but $\hat{p}$ is arbitrary. Given that only $u$ is linked to the trajectory, any transformation in $\{p, \hat{p} \}$ that does not change $u$ describes the same motion. We call \emph{gauge transformation} such transformation that redefines conjugate momentum while keeping kinetic momentum unvaried.}

@@

@@

\bibitem{Shannon} Shannon, C. E.: ``A mathematical theory of communications'', The Bell System Technical Journal, Vol. 27, pp. 379–423, 623–656, (1948)
\bibitem{Jaynes} Jaynes, E. T.: ``Information theory and statistical mechanics'', (1963)
\bibitem{classical_dynamics} J. V. Jose', E. J. Saletan: ``Classical Dynamics'', Cambridge University Press, (1998)
\bibitem{Gromov} Gromov, M. L.: ``Pseudo holomorphic curves in symplectic manifolds''. Inventiones Mathematicae 82: 307–347, (1985)
\bibitem{deGosson} de Gosson, M. A.: ``The symplectic camel and the uncertainty principle: the tip of an iceberg?'', Foundation of Physics 39, pp. 194–214, (2009)
\bibitem{Stewart} Stewart, I.: ``The symplectic camel'', Nature 329(6134), 17–18 (1987)
\bibitem{Lanczos} Lanczos, C.: ``The variational principles of mechanics'', University of Toronto Press (1949)
\bibitem{Synge} Synge, J. L.: Encyclopedia of Physics Vol 3/1, Springer (1960)
\bibitem{Struckmeier} Struckmeier, J.: ``Hamiltonian dynamics on the symplectic extended phase space for autonomous and non-autonomous systems'', J. Phys. A: Math. Gen 38, 1257-1278, (2005)

>>>>

Here we are simply expressing $\omega$, $v$ and $w$ in terms of their respective components and underlining the fact that $\omega$ defines the metric conserved under canonical transformations.

@@

\begin{defn}\label{inertialMass}
Let $p_\alpha$ be a component of conjugate momentum and $u_\alpha \equiv g_{\alpha \beta}u^\beta$ be a contravariant component of the four velocity. We define the \emph{inertial mass} as the cardinality conversion constant $m \equiv \mathfrak{C}^p_u$ between $p$ and $u$. That is, if $dp_\alpha=mdu_\alpha$ then $n^p(dp_\alpha)=n^u(du_\alpha)$.\footnote{Note that the kinematic assumption rules out massless particles: they can have the same position and velocity with different momentum, therefore the same trajectory for different states. When generalizing to field theory, the derivative of the position will be replaced by the derivative of the field (i.e. its oscillation in space-time), which holds for massive and massless fields.}
\end{defn}

@@

Given \ref{initialConditions}, there must exist $p_\alpha=p_\alpha(x^\beta , u^\gamma)$. We express it in terms of $u_\alpha\equiv g_{\alpha \beta} u^\beta$. We have:
\begin{align*}
\omega &= q^i\wedge p_i - ct \wedge E/c \\
&=x^\alpha \wedge p_\alpha \\
&=dx^\alpha \wedge \frac{\partial p_\alpha}{\partial u_\beta}du_\beta + dx^\alpha \wedge \frac{\partial p_\alpha}{\partial x^\gamma}dx^\gamma \\
&=\frac{\partial p_\alpha}{\partial u_\beta}du_\beta dx^\alpha \\
\end{align*}
Consider the expression $m du_\alpha dx^\alpha$: this invariant gives us the density of labels in position and velocity converted to the conjugate variables. That is: $\omega=m du_\alpha dx^\alpha$. Combining the two:
\begin{align*}
\frac{\partial p_\alpha}{\partial u_\beta} &= m \delta^\beta_\alpha \\
p_\alpha &= m g_{\alpha \beta}u^\beta + \hat{p}_\alpha(x)
\end{align*}
where $\hat{p}$ is an arbitrary function.\footnote{The conjugate momentum $p$ and the kinetic momentum $mu$ are therefore different in general: they are linked only through their differentials. Moreover, $p$ is unique only once $\hat{p}$ is fixed, but $\hat{p}$ is arbitrary. Given that $u$ determines the trajectory, any transformation in $\{p, \hat{p} \}$ that does not change $u$ describes the same motion. We call \emph{gauge transformation} such a transformation that redefines conjugate momentum while keeping kinetic momentum unvaried. This is the particle mechanics equivalent of gauge transformations in field theory.}

@@

\section{Acknowledgements}

It is my pleasure to thank Christine A. Aidala, Lydia Bieri, Robert Geroch and Steven J. Miller for helpful discussions and support.

@@

\bibitem{Shannon} Shannon, C. E., ``A mathematical theory of communication'', The Bell System Technical Journal, Vol. 27, pp. 379--423, 623--656, (1948).
\bibitem{Jaynes} Jaynes, E. T., ``Information theory and statistical mechanics'', Statistical Physics 3, pp. 181--218, (1963).
\bibitem{classical_dynamics} J. V. Jos\'{e}, E. J. Saletan, ``Classical Dynamics'', Cambridge University Press, (1998).
\bibitem{Gromov} Gromov, M. L., ``Pseudo holomorphic curves in symplectic manifolds'', Inventiones Mathematicae 82, pp. 307--347, (1985).
\bibitem{deGosson} de Gosson, M. A., ``The symplectic camel and the uncertainty principle: the tip of an iceberg?'', Foundations of Physics 39, pp. 194--214, (2009).
\bibitem{Stewart} Stewart, I., ``The symplectic camel'', Nature 329, pp. 17--18, (1987).
\bibitem{Lanczos} Lanczos, C., ``The variational principles of mechanics'', University of Toronto Press, (1949).
\bibitem{Synge} Synge, J. L., ``Classical dynamics'', Encyclopedia of Physics Vol 3/1, Springer (1960).
\bibitem{Struckmeier} Struckmeier, J., ``Hamiltonian dynamics on the symplectic extended phase space for autonomous and non-autonomous systems'', J. Phys. A: Math. Gen. 38, pp. 1257--1278, (2005).
